Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-24 review_article.Rmd
produces pdf but gives permission errors

# Introduction
<!--
Current best practices: Choropleths
Current exploratory practices: Cartograms: cont, non cont, (tilegrams, Dorlings)
Suggested uses of 
-->

<!-- What data are we working with, why, what are the purposes -->
Cancer statistics represent information on people, usually delivered as an aggregated value for a geopolitical area. Presenting statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone is not sufficient on which to compare different areas, because the populations of different areas are likely different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

@@

A similar method, where each geographic area is represented using a square, tessellated to create a square grid. This method has been used by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR.
Each area is represented by a square of the same dimensions.
Figure \ref{fig:ggcartograms} d shows a hexagon tilegram representing each state using a single square.

@@

```{r ggtilemap, fig.cap = "A tile map and geofaceted map of the Unites States of America. Each state in the tilegram has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a square, each square is the same size. The geofaceted map allows the rates for males and females to be contrasted. It has the same layout as the tilegram, but shows more information per state than ", message=FALSE, warning=FALSE, fig.width=12, fig.height=6, out.width = "100%"}
knitr::include_graphics("figures/gggrids.png")
```

## Geofaceting
Tilegrams can be extended to allow more than just simple colouring of tiles to present data.
@IGF formalises the term geofaceting to describe a grid display, the arrangement of tiles to create a grid that mimics the geographic topology of the set of areas. 
Like tilegrams, the arrangement of the areas can be reused for other spatial distribution visulisations.
Geofaceting have the functionality of facets, often used to replicate visualisations for each subset of the data, in this case the data subsets are geographic.
The amount of information able to be communicated has increased from one value per region in a tilegram to one visualisation, this is a more flexible display.
Virtually any plot display can be fit into the tile representing the areas, allowing displays of multiple variables or values per geographic entity.
As the amount of areas increase care must be taken to ensure legibility of the displays.
@RPIG acknowledges that this display suffers when the areas have very irregular shapes and large size disparities, as manual creation of the grid is still required.

@@

<!-- Alternatives, remove geographic shapes, keep spatial relationship, focus on the outliers/ relationships in population space -->
Alternative maps may replace the unexpected boundaries and shapes with familiar, simple shapes. This makes the spatial distribution the primary concept communicated by the display.
Areas that are geographically close will maintain connectedness in some way, but the population and the statistic will dominate.
Tilegram displays may only allow one variable to be visualised if each region is given consistent map space, this occurs in geofaceting. 
Dorlings provide bi-variate displays as they use the size of the shape for one variable, and colour for another.
These alternatives are especially helpful for data aggregations where administrative boundaries break populations into groups.
The experience of each community may be worth considering, as the experiences shared by the population within them may be similar due to the services and facilities they share.
This display allows a more equitable view of each community, and does not minimise those that operate on a smaller geographic scale.

@@

|  | Choropleth | Contiguous | Non-contiguous | Dorling | Tilegrams |
|:-----------:|------------|-----------|---------------|---------|-----------|
| Preserves shapes | Y | N | Y | N | N |
| Preserves neighbours | Y | Y | Y | S | S |
| Uniform use of shape | N | N | N | Y | Y |
| Bi-variate display^[A bi-variate display refers to a map that allows two variables to be plotted. One variable via the size of the area, and another by filling the area with a colour.] | N | Y | Y | Y | Y |

@@

The following packages were used to create transformations from the sf [@sf] geographical shapes of the states of America [@spData] and the XXX Australian Statistical Areas (Level 3):
The cartogram [@cartogram] package creates contiguous, non-contiguous and Dorling cartograms.
Hexagon tilegram displays were created using the sugarbag package [@sugarbag].

>>>>

# Introduction
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics represent information on people, usually delivered as an aggregated value for a geopolitical area. Presenting statistics requires transforming individual observations into aggregations of communities as geographical units, in large part for privacy protection or political and policy purposes.  The information could be as simple as counts per area (e.g. state, province, local government area, post/zip code). Counts alone is not sufficient on which to compare different areas, because the populations of different areas are likely different. In this case, the counts data needs to be merged with population data to appropriately calibrate it to incidence, for example, rate per 100000 people.  This type of data is collected on a routine basis for public health purposes, and may be made available to the general public as a service to the community. The task, then, is to examine what are the usual ways to communicate cancer statistics, to the public, are there alternative approaches, and what are the pros and cons of these choices.

@@

A similar method, where each geographic area is represented using a square, tessellated to create a square grid. This method has been used by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR.
Each area is represented by a square of the same dimensions.
Figure \ref{fig:ggcartograms} d shows a hexagon tile map representing each state using a single square.

@@

```{r ggtilemap, fig.cap = "A tile map and geofaceted map of the Unites States of America. Each state in the tile map has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a square, each square is the same size. The geofaceted map allows the rates for males and females to be contrasted. It has the same layout as the tile map, but shows more information per state than ", message=FALSE, warning=FALSE, fig.width=12, fig.height=6, out.width = "100%"}
knitr::include_graphics("figures/gggrids.png")
```

## Geofaceting
Tile maps can be extended to allow more than just simple colouring of tiles to present data.
@IGF formalises the term geofaceting to describe a grid display, the arrangement of tiles to create a grid that mimics the geographic topology of the set of areas. 
Like tile maps, the arrangement of the areas can be reused for other spatial distribution visulisations.
Geofaceting have the functionality of facets, often used to replicate visualisations for each subset of the data, in this case the data subsets are geographic.
The amount of information able to be communicated has increased from one value per region in a tile map to one visualisation, this is a more flexible display.
Virtually any plot display can be fit into the tile representing the areas, allowing displays of multiple variables or values per geographic entity.
As the amount of areas increase care must be taken to ensure legibility of the displays.
@RPIG acknowledges that this display suffers when the areas have very irregular shapes and large size disparities, as manual creation of the grid is still required.

@@

<!-- Alternatives, remove geographic shapes, keep spatial relationship, focus on the outliers/ relationships in population space -->
Alternative maps may replace the unexpected boundaries and shapes with familiar, simple shapes. This makes the spatial distribution the primary concept communicated by the display.
Areas that are geographically close will maintain connectedness in some way, but the population and the statistic will dominate.
Tile map displays may only allow one variable to be visualised if each region is given consistent map space, this occurs in geofaceting. 
Dorlings provide bi-variate displays as they use the size of the shape for one variable, and colour for another.
These alternatives are especially helpful for data aggregations where administrative boundaries break populations into groups.
The experience of each community may be worth considering, as the experiences shared by the population within them may be similar due to the services and facilities they share.
This display allows a more equitable view of each community, and does not minimise those that operate on a smaller geographic scale.

@@

|  | Choropleth | Contiguous | Non-contiguous | Dorling | Tile maps |
|:-----------:|------------|-----------|---------------|---------|-----------|
| Preserves shapes | Y | N | Y | N | N |
| Preserves neighbours | Y | Y | Y | S | S |
| Uniform use of shape | N | N | N | Y | Y |
| Bi-variate display^[A bi-variate display refers to a map that allows two variables to be plotted. One variable via the size of the area, and another by filling the area with a colour.] | N | Y | Y | Y | Y |

@@

The following packages were used to create transformations from the sf [@sf] geographical shapes of the states of America [@spData] and the Australian Statistical Areas (Level 3):
The cartogram [@cartogram] package creates contiguous, non-contiguous and Dorling cartograms.
Hexagon tile map displays were created using the sugarbag package [@sugarbag].
